This practical aims to guide you through some of the key ideas in
**ggplot2**. As with the first practical, feel free to experiment. Some of the functions introduced in this practical haven't been explicitly covered in the notes. Use the built-in R help or the **ggplot2** help pages at

> http://had.co.nz/ggplot2/

\noindent as needed.


## Over plotting

Scatter plots are very useful. However, when we have a large data set, points
will be plotted on top of each other obscuring the relationship. We call this
problem over plotting. There are a few techniques we can use to help, although
the best solution is often problem specific.

To begin with we will create an example data frame:

```{r,overplotting, tidy=FALSE, echo=1:4}
## If your computer is slow when plotting reduce the value of n
library("jrGgplot2Bio")
library("ggplot2")
df = overplot_data(n = 20000)
h = ggplot(df) + geom_point(aes(x, y))
```

```{r,echo=FALSE, fig.margin = TRUE, fig.cap="A scatter plot that suffers from over plotting.", fig.height = 4}
h
```

\noindent We can create a simple scatter plot of this data using the following command

```{r, echo = TRUE}
h = ggplot(df) + geom_point(aes(x, y))
```

\noindent This plot isn't particularly good. Try to improve it by using a
combination of:


- changing the transparency level: `alpha`;^[`alpha` takes a value between $0$ and $1$.]
- change the shape: `shape=1` and `shape=`.'`
- use some jittering - `geom_jitter`.
- adding a contour to the plot using `stat_density2d`.
- What does

```{r,cache=TRUE, fig.keep='none', tidy=FALSE, echo = TRUE}
h +  stat_density2d(aes(x, y, fill = ..density..),
                contour = FALSE, geom = "tile")
```
do?    
- What does `stat_bin2d()` and `stat_binhex()` do - add it to the
  plot to find out! Try varying the parameters `bins` and `binwidth`.


## Displaying distributions

Lets return to the outbreaks data to look at how to visualise distributions. 

```{r, echo = TRUE}
data(outbreaks, package = "jrGgplot2Bio")
```

\noindent First lets calculate a new variable, log transformed illnesses. This is because the distribution of illnesses is very skewed, recall some of the plots from the first practical.

```{r,tidy=FALSE, echo = TRUE}
outbreaks$log_illness = log(outbreaks$illnesses)
```

\noindent We can construct a histogram of log transformed illness counts using the following commands:

```{r,fig.keep='none', message=FALSE, echo = TRUE}
ggplot(data = outbreaks) +
  geom_histogram(aes(x = log_illness))
```

```{r,echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Histogram of log illnesses caused by three pathogens in the USA", fig.height = 4}
ggplot(data = outbreaks) +
  geom_histogram(aes(x = log_illness))
```


\noindent to get figure 2. Let's experiment a bit.

1. Change the `bins` in the `geom_histogram`. What value do you
  think is best?
2. What happens when you set `fill=species` in the `geom_histogram`
  aesthetic? What other options can you change?^[Look at the `geom_histogram` help page: http://had.co.nz/ggplot2/geom_histogram.html}
3. Try `geom_density`. Set `fill=species` and change the `alpha` value.



## Copy cat

```{r,F1, echo=FALSE, message=FALSE, fig.keep='none',results='hide'}
library(dplyr)
data(yeast, package = "jrGgplot2Bio")
yeast = yeast %>% filter(class == "CYT" | class == "EXC")
```

The aim of this section is to recreate the graphics in figure 3. Feel
free to experiment. To begin, load the package and subset the data.

```{r,results='hide', echo = TRUE}
library(dplyr)
data(yeast, package = "jrGgplot2Bio")
yeast = yeast %>% filter(class == "CYT" | class == "EXC")
```

```{r,dev='png', out.width='\\textwidth', echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Two measure of signal sequence recognition and two groups of proteins", fig.height = 4}
ggplot(yeast, aes(x = gvh, y = mcg)) +
  geom_point(aes(col = class)) +
  geom_density2d(aes(col = class), alpha = 0.5) +
  xlab("von Heijne's method") +
  ylab("McGeoch's method")
```

1. Figure 3: Create a scatter plot of McGeoch's method for signal sequence recognition,
  `mcg`, against von Heijne's method, `gvh`. To get started:

```{r,fig.keep='none', tidy=FALSE, echo = TRUE}
ggplot(yeast, aes(x = gvh, y = mcg)) +
  geom_point() +
  xlab("von Heijne's method")
```

Now add the densisity lines and change the $y$-axis label.
Hint: try `geom_density2d` and `ylab(`New label')`.

```{r, fig.keep='none'}
ggplot(yeast, aes(x = gvh, y = mcg)) +
  geom_point() +
  geom_density2d() +
  xlab("von Heijne's method") +
  ylab("McGeoch's method") 
```

2. Using the `col` aesthetic, add colour to the points and density lines dependent upon the `class` variable. For the final touch adjust the transparency of the density lines.
```{r, fig.keep='none'}
ggplot(yeast, aes(x = gvh, y = mcg)) +
  geom_point(aes(col = class)) +
  geom_density2d(aes(col = class), alpha = 0.5) +
  xlab("von Heijne's method") +
  ylab("McGeoch's method")
```

## Solutions

Solutions are contained within this package:
```{r,eval=FALSE, echo=TRUE}
vignette("solutions2", package = "jrGgplot2Bio")
```
